The microbiota of the proximal and distal human colon

### 2017-08-16 submission/manuscript.Rmd
some Pat edits made

---
title: Unprepared colonoscopy identifies microbiota specific to the proximal and distal
  human colon
csl: mbio.csl
fontsize: 11pt
output:
  pdf_document:
    includes: null
    keep_tex: yes
  html_document: default
  word_document: default
header-includes:
- \usepackage{setspace}
- \doublespacing
- \usepackage{lineno}
- \linenumbers
geometry: margin=1.0in
bibliography: references.bib
---
\vspace{35mm}

Running title: Unprepared colonoscopy identifies proximal and distal human colonic microbiota

@@

Kaitlin J. Flynn^1^, Charles C. Koumpouras^1^, Mack T. Ruffin IV^2^, Danielle Kimberly Turgeon^3^, and Patrick D. Schloss^1$\dagger$^

@@

$\dagger$ Corresponding author: pschloss@umich.edu

@@

The human colon contains chemicals and nutrients that change along the proximal to distal gut axis. These gradients create microenvironments that affect the distrubution and composition of the gut microbiota. The microbiome has been implicated in the colonic diseases colorectal cancer (CRC) and inflammatory bowel disease (IBD). Further, these diseases exhibit different symptoms depending on the location of the colon they are found in. CRC tumors of the proximal and distal colon are morphologically and genetically distinct. Similarly, inflammatory bowel diseases such as Crohn's are typically exacerbated in the proximal instestine while ulcerative colitis patients often experience symptoms in the distal colon. Previous analysis of the fecal microbiota from healthy and CRC or IBD patients has revealed different microbial signatures associated with these diseases. We extended these observations of the fecal microbiome to include analysis of the proximal and distal healthy human colon. We used a two-colonoscope approach on subjects that had not undergone standard bowel preparation procedure. This technique allowed us to characterize the native proximal and distal luminal and mucosal microbiome without prior chemical disruption. 16S rRNA gene sequencing was performed on proximal and distal mucosal and luminal biopsies and home-collected stool for 20 healthy individuals. Diversity analysis revealed that each site contained a diverse community and that a patient’s samples were more similar to each other than to that of other individuals. Comparison of all samples to fecal samples taken at exit uncovered that the feces were most similar to samples taken from the distal lumen, likely reflecting the anatomical structure of the colon. Since we could not differentiate sites along the colon based on community structure or community membership alone, we employed the Random Forest machine-learning algorithm to identify key species that distinguish biogeographical sites. Random Forest classification models were built using taxa abundance and sample location and revealed distinct populations that were found in each location. A model differentiating the proximal mucosa and lumen was built with an AUC of `r signif(cv10f_roc_proximal_bs$auc, digits=3)`. The proximal mucosa had a higher abundance of the genera _Enterobacteriaceae_ and _Bacteriodes_. _Peptoniphilus, Anaerococcus, Finegoldia,_ and _Turicibacter_ were most likely to be found in distal mucosal samples versus distal luminal samples (AUC = `r signif(cv10f_roc_distal_bs$auc, digits=3)`). The classification model performed well (AUC = `r signif(cv10f_roc_muc$auc, digits=3)`) when classifying mucosal samples into proximal or distal sides, but separating luminal samples from each side proved more challenging (AUC = `r signif(cv10f_roc_lum$auc, digits =3)`). The distal mucosa was found to have high populations of _Finegoldia, Murdochiella_ and _Porphyromonas_. Proximal and distal luminal samples were comprised of many of the same taxa, likely reflecting the fact that stool moves along the colon from the proximal to distal end. By sampling the unprepped human colon, our results have identified distinct bacterial populations native to the proximal and distal sides. Further investigation of these bacteria may elucidate if and how these groups contribute to different pathogenesis processes on the respective sides of the colon.

@@

The human colon is an ecosystem comprised of several different microenvironments inhabited by resident bacterial members of the microbiome. Concentrations of oxygen, water and anti-microbial peptides change along the gut axis and influence what populations of microbes reside in each location. Microenvironments differ not only longitutinally along the colon, but latitudinally from the epithelium to mucosa to intestinal lumen, offering several sites for different microbial communities to flourish. The identity of these specific microbes and communities are important for understanding the etiology of complex colon diseases such as Inflammatory Bowel Disease (IBD) and Colorectal Cancer (CRC). IBD and CRC are known to be preceded or accelerated by perturbations in gut microbes @Yamauchi2012, @Forbes2016, @Halfvarson2017. The severity, symptoms, morbidity and mortality of these diseases is known to vary based upon the biogeographical location in which they occur. For instance, CRC tumors that arise on the distal side of the colon are infiltrating lesions that present with painful symptoms @Benedix2010. In contrast, 47% of CRCs are caused by proximal-sided colon tumors that are sessile and form along the wall of the colon, often remaining asymptomatic until advanced carcinogenesis @Benedix2010. The distal and proximal sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively, in addition to variation in the previously mentioned chemical gradients @Yamauchi2012, @Albenberg2014, @Donaldson2015. In IBD patients, disease flares in the distal colon are usually indicative of ulcerative colitis (UC), whereas Crohn's disease (CD) patients typically experience disease in the small intestine, ileum and proximal colon @Forbes2016. UC presents as large and highly inflammed mucosal ulcers, where as CD lesions are often smaller and have areas of normal tissue distributed amongst the flare @Forbes2016. Thus, given the varied physiology of the proximal-distal axis of the colon and known differences in disease patterns at these sites, symbiotic microbes and their metabolites likely vary as well, and may influence the heterogenous disease prognoses of IBD and CRC. Because CRC is a long-term complication of IBD, the distribution of microbes is important to understanding the pathophysiology of both diseases.

Several recent findings have shown that development and progression of IBD or CRC can be attributed to specific molecular events as a result of interactions between the gut microbiota and human host @Yamauchi2012, @Halfvarson2017. For instance, comparison of the bacteria present on CRC tumors with those found on nearby healthy tissue has identified specific species that are tumor-associated @Dejea2014. These species include the oral pathogens _Fusobacterium nucleatum_ and _Porphyromonas asacharolytica_. Interestingly, these periodontal pathogens have been highly predictive of whether a patient had CRC tumors or not in our prior human stool classification studies @Baxter2016. _F. nucleatum_ has also been found to be elevated in the stool and biopsies of patients with IBD as compared to healthy controls @Strauss2011. Furthermore, studies of _F. nucleatum_ isolated from mucosal biopsies showed that more invasive  _F. nucleatum_ positively correlates with IBD disease level @Strauss2011. Like many intestinal pathogens, the bacteria appear to have a high-impact despite being lowly-abundant in the community @Forbes2016. The molecular capabilities of these rare taxa may contribute to the colonic disease state. These studies often examined only shed human stool or the small intestine, preventing fine-resolution analysis of paired samples from the proximal and distal sides of the colon. Similarly, comparisons of on- or off-tumor/lesion bacteria rarely have matched tissue from the other side of the colon from the same patient, limiting what conclusions can be drawn about the colonic microbiome overall, let alone at that specific site. Due to these limitations, the contribution of the gut microbiota to IBD and CRC disease location in the colon is largely undefined. Characterizing these communities could provide needed insight into disease etiology, including how the disruption of the healthy community could promote the initiation or proliferation of the distinct proximal and distal CRC tumors or IBD flares.

The few existing profiles of the microbial biogeography of the gut have been limited by sample collection methods. The majority of human gut microbiome studies have been performed on whole shed stool or on samples collected during colonoscopy procedures. While the latter method allows investigators to acquire samples from inside the human colon, typically this procedure is preceded by the use of bowel preparation methods such as the consumption of laxatives to cleanse the bowel. Bowel preparation is essential for detecting cancerous or precancerous lesions in the colon, but complicates microbiome profiling as the chemicals strip the bowel of contents and disrupt the mucosal layer @Jalanka2014, @Harrell2012. As such, what little information we do have about the biogeographical distribution of the microbes in the proximal and distal colon is confounded by the bowel preparation procedure.

>>>>

---
title: The microbiota of the proximal and distal human colon
csl: mbio.csl
fontsize: 11pt
output:
  pdf_document:
    includes: null
    keep_tex: yes
  html_document: default
  word_document: default
header-includes:
- \usepackage{setspace}
- \doublespacing
- \usepackage{lineno}
- \linenumbers
geometry: margin=1.0in
bibliography: references.bib
---
\vspace{35mm}

Running title: The microbiome of the proximal and distal human colon

@@

Kaitlin J. Flynn^1^, Charles C. Koumpouras^1^, Mack T. Ruffin IV^2^, Danielle Kimberly Turgeon^3$\dagger$^, and Patrick D. Schloss^1$\dagger$^

@@

$\dagger$ Corresponding authors: kturgeon@med.umich.edu and pschloss@umich.edu

@@

Chemical and nutrient gradients along the human colon create microenvironments that affect the distrubution and composition of the gut microbiota. The microbiome has been implicated in colorectal cancer (CRC) and inflammatory bowel disease (IBD). Further, these diseases exhibit different symptoms depending on the location of the colon they are found in. CRC tumors of the proximal and distal colon are morphologically and genetically distinct. Similarly, inflammatory bowel diseases such as Crohn's are typically exacerbated in the proximal instestine while ulcerative colitis patients often experience symptoms in the distal colon. Previous analysis of the fecal microbiota from healthy and CRC or IBD patients has revealed different microbial signatures associated with these diseases. We extended these observations of the fecal microbiome to include analysis of the proximal and distal healthy human colon. We used a two-colonoscope approach on subjects that had not undergone standard bowel preparation procedure. In contrast to previous efforts using prepared colonoscopy, this technique allowed us to characterize the native proximal and distal luminal and mucosal microbiome without prior chemical disruption. 16S rRNA gene sequencing was performed on proximal and distal mucosal and luminal biopsies and home-collected stool for 20 healthy individuals. Diversity analysis revealed that each site contained a diverse community and that a patient’s samples were more similar to each other than to that of other individuals. A patient's feces were most similar to samples taken from the distal lumen, likely reflecting the anatomical structure of the colon. Since we could not differentiate the communities in the proximal and distal colon based on community structure or community membership alone, we employed the Random Forest machine-learning algorithm to identify key taxa that the two sites in the lumen and mucosa. Random Forest classification models were built using taxa abundance and sample location and revealed distinct populations that were found in each location. A model differentiating the proximal mucosa and lumen was built with an AUC of `r signif(cv10f_roc_proximal_bs$auc, digits=3)`. The proximal mucosa had a higher abundance of the genera _Enterobacteriaceae_ and _Bacteriodes_. _Peptoniphilus, Anaerococcus, Finegoldia,_ and _Turicibacter_ were most likely to be found in distal mucosal samples versus distal luminal samples (AUC = `r signif(cv10f_roc_distal_bs$auc, digits=3)`). The classification model performed well (AUC = `r signif(cv10f_roc_muc$auc, digits=3)`) when classifying mucosal samples into proximal or distal sides, but separating luminal samples from each side proved more challenging (AUC = `r signif(cv10f_roc_lum$auc, digits =3)`). The distal mucosa was found to have high populations of _Finegoldia, Murdochiella_ and _Porphyromonas_. Proximal and distal luminal samples were comprised of many of the same taxa, likely reflecting the fact that stool moves along the colon from the proximal to distal end. By sampling the unprepped human colon, our results have identified distinct bacterial populations native to the proximal and distal sides. Further investigation of these bacteria may elucidate if and how these groups contribute to different pathogenesis processes on the respective sides of the colon.

#### wordcount?

@@

The human colon is an ecosystem comprised of several different microenvironments inhabited by resident bacterial members of the microbiome. Concentrations of oxygen, water and anti-microbial peptides change along the gut axis and influence what populations of microbes reside in each location. Microenvironments differ not only longitutinally along the colon, but latitudinally from the epithelium to mucosa to intestinal lumen, offering several sites for different microbial communities to flourish. The identity of these specific microbes and communities are important for understanding the etiology of complex colon diseases such as Inflammatory Bowel Disease (IBD) and Colorectal Cancer (CRC). IBD and CRC are known to be preceded or accelerated by perturbations in gut microbes @Yamauchi2012, @Forbes2016, @Halfvarson2017. The severity, symptoms, morbidity and mortality of these diseases is known to vary based upon the biogeographical location in which they occur. For instance, CRC tumors that arise on the distal side of the colon are infiltrating lesions that present with painful symptoms @Benedix2010. In contrast, 47% of CRCs are caused by proximal-sided colon tumors that are sessile and form along the wall of the colon, often remaining asymptomatic until advanced carcinogenesis @Benedix2010. The distal and proximal sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively, in addition to variation in the previously mentioned chemical gradients @Yamauchi2012, @Albenberg2014, @Donaldson2015. In IBD patients, disease flares in the distal colon are usually indicative of ulcerative colitis (UC), whereas Crohn's disease (CD) patients typically experience disease in the small intestine, ileum and proximal colon @Forbes2016. UC presents as large and highly inflammed mucosal ulcers, where as CD lesions are often smaller and have areas of normal tissue distributed amongst the flare @Forbes2016. Thus, given the varied physiology of the proximal-distal axis of the colon and known differences in disease patterns at these sites, symbiotic microbes and their metabolites likely vary as well, and may influence the heterogenous disease prognoses of IBD and CRC. Because CRC can be a long-term complication of IBD, the distribution of microbes is important to understanding the pathophysiology of both diseases.

Several recent findings have shown that development and progression of IBD or CRC can be attributed to specific molecular events as a result of interactions between the gut microbiota and human host @Yamauchi2012 @Halfvarson2017. For instance, comparison of the bacteria present on CRC tumors with those found on nearby healthy tissue has identified specific species that are tumor-associated @Dejea2014. These species include the oral pathogens _Fusobacterium nucleatum_ and _Porphyromonas asacharolytica_. Interestingly, these periodontal pathogens have been highly predictive of whether a patient had CRC tumors or not in our prior human stool classification studies @Baxter2016. _F. nucleatum_ has also been found to be elevated in the stool and biopsies of patients with IBD as compared to healthy controls @Strauss2011. Furthermore, studies of _F. nucleatum_ isolated from mucosal biopsies showed that more invasive  _F. nucleatum_ positively correlates with IBD disease level @Strauss2011. Like many intestinal pathogens, the bacteria appear to have a high-impact despite being lowly-abundant in the community @Forbes2016. The physiology of these rare taxa may contribute to the colonic disease state. These studies often examined only shed human stool or the small intestine, preventing fine-resolution analysis of paired samples from the proximal and distal sides of the colon. Similarly, comparisons of on- or off-tumor/lesion bacteria rarely have matched tissue from the other side of the colon from the same patient, limiting what conclusions can be drawn about the colonic microbiome overall, let alone at that specific site. Due to these limitations, the contribution of the gut microbiota to IBD and CRC disease location in the colon is largely undefined. Characterizing these communities could provide needed insight into disease etiology, including how the disruption of the healthy community could promote the initiation or proliferation of the distinct proximal and distal CRC tumors or IBD flares.

The few existing profiles of the microbial biogeography of the gut have been limited by sample collection methods. The majority of human gut microbiome studies have been performed on whole shed feces or on samples collected during colonoscopy procedures. While the latter method allows investigators to acquire samples from inside the human colon, typically this procedure is preceded by the use of bowel preparation methods such as the consumption of laxatives to cleanse the bowel. Bowel preparation is essential for detecting cancerous or precancerous lesions in the colon, but complicates microbiome profiling as the chemicals strip the bowel of contents and disrupt the mucosal layer @Jalanka2014, @Harrell2012. As such, what little information we do have about the biogeographical distribution of the microbes in the proximal and distal colon is confounded by the bowel preparation procedure.
